Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-08-12 ms.tex
Added description and figure for Test 11.  Added 1/dt factor for rad. pressure.  Removed Test 12.

  We describe a photon-conserving numerical method that solves the
  radiative transfer equation, using a spatially-adaptive ray tracing
  scheme, and its parallel implementation into the adaptive mesh
  refinement (AMR) code, Enzo.  Coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive timestepping scheme and a
  fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  This method linearly scales
  with the number of point sources and number of grid cells.  To
  combat this, we use a novel method of merging rays at large radius,
  using a tree method, which we briefly describe.  Our implementation
  is scalable to O(10$^3$) processors on distributed and shared memory
  machines and can include radiation pressure and secondary
  ionizations from X-ray radiation.  Our implementation is included in
  the newest public release of Enzo.

@@

A ray propagates and splits until 
%
\begin{enumerate}
\item its photon flux is almost fully absorbed ($>99.9\%$) in a single
  cell, which significantly reduces the computational time if the
  radiation volume filling fraction is small,
\item the photon has travelled $c \times dt_P$, where $dt_P$ is the
  radiative transfer timestep,
\item the photon leaves the computational domain with isolated
  boundary conditions, or
\item the photon travels $\sqrt{3}$ of the simulation box length with
  periodic boundary conditions.
\end{enumerate}
%
In the second case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  In this sense, our method retains the time
derivative of the radiative transfer equation.

@@

The absorption of radiation transfers momentum from photons to the
absorbing medium, i.e. radiation pressure.  This is easily computed by
considering the momentum
\begin{equation}
  d\mathbf{p}_\gamma = \frac{dP \; E_{\rm ph}}{c} \; \hat{r}
\end{equation}
of the absorbed radiation from the incoming ray, where $\hat{r}$ is
the normal direction of the ray.  The resulting acceleration of the
gas due to radiation pressure is 
\begin{equation}
  d\mathbf{a} = d\mathbf{p}_\gamma / (\rho \; V_{\rm cell}),
\end{equation}
where $\rho$ is the gas density inside the cell.  This acceleration is
then added to the other forces, e.g. gravity and gas pressure, in the
calculation.

@@

@@

We use a domain that is 2 kpc wide and has an ambient medium with
$\rho_0 = 10^{-3}$ \cubecm, $T = 10^4$ K, $x_e = 0.99$.  We place a
dense clump with $\rho/\rho_0 = 100$, $T = 100$ K, $x_e = 10^{-3}$,
and $r = 250$ pc, at the center of the box.  Radiation is emitted in
two polar cones with $\theta_c = \pi/6$ with a total luminosity of
$10^{49}$ erg s$^{-1}$ and a 17.6 eV mono-chromatic spectrum.  This
results in $t_{\rm rec} = 1.22$ Myr and $R_s = 315$ pc, just outside
of the sphere.  The base grid has a resolution of 64$^3$, and it is
refined with the same overdensity criterion as Test 8.

@@

\subsection{Test 12: \ion{H}{2} region in a static potential}

Next we construct a problem that is similar to the Rayleigh-Taylor
test problem (Test 9) in nature but is spherically symmetric.  We use
the same physical setup as Test 5 with a uniform medium and the
radiation source at the origin.  It has a luminosity of $5 \times
10^{48}$ ph s$^{-1}$ and a mono-chromatic spectrum of 14.6 eV.  We
choose this energy so the ionization front is relatively sharp but
still heats the gas to $10^4$ K.  The ambient medium has a density of
$10^{-3}~\cubecm$ and an initial temperature of 100 K.  A point source
of gravity also resides at the origin.  To have an equal gravitational
acceleration $g(r) = g_0 / r^3$ at the Str\"{o}mgren radius $R_s =
5.4$ kpc ($r = 0.36$) as Test 9, we set the strength of the point
source $g_0 = 4.67 \times 10^{-3}$.  The gravity point source has the
same effect on any pertubations as the constant gravitational force in
Rayleigh-Taylor instability test.

\subsection{Test 13: Time variations of the source luminosity}

>>>>

  We describe a photon-conserving numerical method that solves the
  radiative transfer equation, using a spatially-adaptive ray tracing
  scheme, and its parallel implementation into the adaptive mesh
  refinement (AMR) code, Enzo.  Coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive timestepping scheme and a
  fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  This method linearly scales
  with the number of point sources and number of grid cells.  To
  combat this, we use a novel method of merging rays at large radius,
  using a tree method, which we briefly describe.  Our implementation
  is scalable to O(10$^3$) processors on distributed and shared memory
  machines and can include radiation pressure and secondary
  ionizations from X-ray radiation.  It is included in the newest
  public release of Enzo.

@@

A ray propagates and splits until 
%
\begin{enumerate}
\item the photon has travelled $c \times dt_P$, where $dt_P$ is the
  radiative transfer timestep,
\item its photon flux is almost fully absorbed ($>99.9\%$) in a single
  cell, which significantly reduces the computational time if the
  radiation volume filling fraction is small,
\item the photon leaves the computational domain with isolated
  boundary conditions, or
\item the photon travels $\sqrt{3}$ of the simulation box length with
  periodic boundary conditions.
\end{enumerate}
%
In the second case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  In this sense, our method retains the time
derivative of the radiative transfer equation.

@@

The absorption of radiation transfers momentum from photons to the
absorbing medium, i.e. radiation pressure.  This is easily computed by
considering the momentum
\begin{equation}
  d\mathbf{p}_\gamma = \frac{dP \; E_{\rm ph}}{c\;dt} \; \hat{r}
\end{equation}
of the absorbed radiation from the incoming ray, where $\hat{r}$ is
the normal direction of the ray and $dt$ is the radiative transfer
timestep.  The resulting acceleration of the gas due to radiation
pressure is
\begin{equation}
  d\mathbf{a} = d\mathbf{p}_\gamma / (\rho \; V_{\rm cell}),
\end{equation}
where $\rho$ is the gas density inside the cell.  This acceleration is
then added to the other forces, e.g. gravity and gas pressure, in the
calculation.

@@

\begin{figure}[t]
  \epsscale{1}
  \plotone{fig/test11_slices}
  \caption{\label{fig:test11} Test 11 (Collimated radiation from a
    dense clump).  Slices of density (top) and temperature (bottom) at
    $t = 0.1, 3.25, 10.75, 23.25$ Myr.  The conical \ion{H}{2} region
    drives shocks transversely into the overdense sphere and creates
    polar champange flows.  The ambient medium is heated to $T \sim 3
    \times 10^4$ K as the ionization front passes the
    constant-pressure cloud surface.  The ionization front changes
    from D-type to R-type after it enters the ambient medium.}
  \epsscale{1}
\end{figure}

@@

We use a domain that is 2 kpc wide and has an ambient medium with
$\rho_0 = 10^{-3}$ \cubecm, $T = 10^4$ K, $x_e = 0.99$.  We place a
dense clump with $\rho/\rho_0 = 100$, $T = 100$ K, $x_e = 10^{-3}$,
and $r = 250$ pc, at the center of the box.  Radiation is emitted in
two polar cones with $\theta_c = \pi/6$ with a total luminosity of
$10^{49}$ erg s$^{-1}$ and a 17.6 eV mono-chromatic spectrum.  This
results in $t_{\rm rec} = 1.22$ Myr and $R_s = 315$ pc, just outside
of the sphere.  The base grid has a resolution of 64$^3$, and it is
refined with the same overdensity criterion as Test 8.  We run this
test for 25 Myr.

We illustrate the expansion of the \ion{H}{2} region created by the
beamed radiation in Figure \ref{fig:test11}.  Before 3 Myr, the
\ion{H}{2} region is conical and contained within the dense clump,
depicted in the $t = 0.1$ Myr snapshot of the system.  At this time,
the ionization front is transitioning from R-type to D-type in the
transverse direction of the cone.  This can be seen in the minute
overdensities on the \ion{H}{2} transverse surface.  When it breaks
out of the overdensity, it creates a champange flow, where the
ionization front transitions back to a weak R-type front.  The cloud
surface is a constant-pressure contact discontinuity (CD) with a
density jump of 100.  After the front heats the gas at the CD, there
exists a pressure difference of $\sim 100$.  In response, the high
density gas accelerates into the ambient medium and heats it to $3
\times 10^4$ K.  Additionally a rarefaction wave travels towards the
clump center.  At later times, the transverse D-type front continues
through the clump, eventually creating a disk-like structure at the
final time.  The polar champange flows proceed to flow outwards,
creating a density shell with a diffuse ($10^{-28}$ \cubecm) and warm
(5000 K) medium in its wake.

@@

% \subsection{Test 12: \ion{H}{2} region in a static potential}

% Next we construct a problem that is similar to the Rayleigh-Taylor
% test problem (Test 9) in nature but is spherically symmetric.  We use
% the same physical setup as Test 5 with a uniform medium and the
% radiation source at the origin.  It has a luminosity of $5 \times
% 10^{48}$ ph s$^{-1}$ and a mono-chromatic spectrum of 14.6 eV.  We
% choose this energy so the ionization front is relatively sharp but
% still heats the gas to $10^4$ K.  The ambient medium has a density of
% $10^{-3}~\cubecm$ and an initial temperature of 100 K.  A point source
% of gravity also resides at the origin.  To have an equal gravitational
% acceleration $g(r) = g_0 / r^3$ at the Str\"{o}mgren radius $R_s =
% 5.4$ kpc ($r = 0.36$) as Test 9, we set the strength of the point
% source $g_0 = 4.67 \times 10^{-3}$.  The gravity point source has the
% same effect on any pertubations as the constant gravitational force in
% Rayleigh-Taylor instability test.

\subsection{Test 12: Time variations of the source luminosity}
